---
title: "DataFest 2016 <br/> Data visualization with R, ggplot2, and Shiny"
author: "Mine Çetinkaya-Rundel"
date: "March 28, 2016"
output: ioslides_presentation
runtime: shiny
---

```{r, echo=FALSE, results='hide'}
options(width = 90)

suppressMessages(library(dplyr))
suppressMessages(library(ggplot2))
suppressMessages(library(shiny))
```


# Set up & Configuration

## Datasets, presentation, source code

- That data we will be using today is available at

    - http://bit.ly/df2016_viz_data

Download this file onto your computer. The codebook can also be found here.

- All source files can be found at

    - http://bit.ly/df2016_viz

## Load data

```{r}
movies <- read.csv("data/movies.csv", stringsAsFactors = FALSE)
```

# ggplot2

## ggplot2

- ggplot2 is a plotting system for R, based on the grammar of graphics, which tries to take the good parts of base and lattice graphics and none of the bad parts. 

- It takes care of many of the fiddly details that make plotting a hassle (like drawing legends) as well as providing a powerful model of graphics that makes it easy to produce complex multi-layered graphics.

<br>

Source: http://ggplot2.org/

## Why ggplot2?

- **Statistical:** Visualize $Y \sim X | Z$ with ease

- **Technical:** Human readable syntax

- **Bonus:** Aesthetically pleasing

## With ggplot2 {.smaller}

```{r fig.height = 4}
ggplot(data = movies, aes(x = critics_score, y = audience_score, color = title_type)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)
```

## With base R {.smaller}

```{r eval = FALSE}
plot(y = movies$audience_score, x = movies$critics_score, 
     col = factor(movies$title_type))

doc <- movies[movies$title_type == "Documentary", ]
ff <- movies[movies$title_type == "Feature Film", ]
tv <- movies[movies$title_type == "TV Movie", ]

m_doc <- lm(audience_score ~ critics_score, data = doc)
m_ff <-  lm(audience_score ~ critics_score, data = ff)
m_tv <-  lm(audience_score ~ critics_score, data = tv)

abline(m_doc, col = 1)
abline(m_ff, col = 2)
abline(m_tv, col = 3)

legend("topleft", levels(factor(movies$title_type)), 
       col = c(1,2,3), lty = 1)
```


## With base R {.smaller}

```{r fig.height = 4, echo = FALSE}
plot(y = movies$audience_score, x = movies$critics_score, 
     col = factor(movies$title_type))

doc <- movies[movies$title_type == "Documentary", ]
ff <- movies[movies$title_type == "Feature Film", ]
tv <- movies[movies$title_type == "TV Movie", ]

m_doc <- lm(audience_score ~ critics_score, data = doc)
m_ff <-  lm(audience_score ~ critics_score, data = ff)
m_tv <-  lm(audience_score ~ critics_score, data = tv)

abline(m_doc, col = 1)
abline(m_ff, col = 2)
abline(m_tv, col = 3)

legend("topleft", levels(factor(movies$title_type)), 
       col = c(1,2,3), lty = 1)
```

# Grammar of Graphics

## The Grammar of Graphics

- Visualisation concept created by Wilkinson (1999)
    - to define the basic elements of a statistical graphic 

- Adapted for R by Wickham (2009) who created the `ggplot2` package
    - consistent and compact syntax to describe statistical graphics
    - highly modular as it breaks up graphs into semantic components 

- Is not a guide which graph to choose and how to convey information best!

<br>

Source: https://rpubs.com/timwinke/ggplot2workshop

## The Grammar of Graphics - Terminology

A statistical graphic is a...

- mapping of **data**
- to **aesthetic attributes** (color, size, xy-position)
- using **geometric objects** (points, lines, bars)
- with data being **statistically transformed** (summarised, log-transformed)
- and mapped onto a specific **facet** and **coordinate system**

## Audience score vs. critics score {.smaller}

```{r echo=FALSE, fig.height=3, fig.width=5, warning=FALSE}
ggplot(data = movies, aes(x = audience_score, y = critics_score)) +
  geom_point()
```

- Which data is used as an input?
- What geometric objects are chosen for visualization?
- What variables are mapped onto which attributes?
- What type of scales are used to map data to aesthetics?
- Are the variables statistically transformed before plotting?

## Audience score vs. critics score - code {.smaller}

```{r fig.height=3, fig.width=5}
ggplot(data = movies, aes(x = audience_score, y = critics_score)) +
  geom_point()
```

## Altering features {.smaller}

```{r echo=FALSE, fig.height=3, fig.width=5, warning=FALSE}
ggplot(data = movies, aes(x = audience_score, y = critics_score)) +
  geom_point(alpha = 0.5, color = "blue")
```

- How did the plot change?
- Are these changes based on data (i.e. can be mapped to variables in the dataset) or
changes in preferences for geometric objects?

## Altering features - code

```{r fig.height=3, fig.width=5}
ggplot(data = movies, aes(x = audience_score, y = critics_score)) +
  geom_point(alpha = 0.5, color = "blue")
```

## Faceting {.smaller}

```{r echo=FALSE, fig.height=3, fig.width=8, warning=FALSE}
ggplot(data = movies, aes(x = audience_score, y = critics_score, color = genre)) +
  geom_point(alpha = 0.5) +
  facet_grid(. ~ title_type)
```

- How did the plot change?
- Are these changes based on data (i.e. can be mapped to variables in the dataset) or
changes in preferences for geometric objects?

## Faceting - code {.smaller}

```{r fig.height=3, fig.width=8, warning=FALSE}
ggplot(data = movies, aes(x = audience_score, y = critics_score, color = genre)) +
  geom_point(alpha = 0.5) +
  facet_grid(. ~ title_type)
```

## More faceting {.smaller}

```{r fig.height=4, fig.width=8, warning=FALSE}
ggplot(data = movies, aes(x = audience_score, y = critics_score, color = genre)) +
  geom_point(alpha = 0.5) +
  facet_grid(audience_rating ~ title_type)
```

## Anatomy of ggplots

```
ggplot(data = [dataframe], aes(x = [var_x], y = [var_y], 
       color = [var_for_color], fill = [var_for_fill], 
       shape = [var_for_shape])) +
    geom_[some_geom] +
    ... # other options
```

# Various plots

## Histograms

```{r fig.height=3, fig.width=7, warning=FALSE}
ggplot(data = movies, aes(x = audience_score)) +
  geom_histogram(binwidth = 5)
```

## Boxplots

```{r fig.height=3, fig.width=8, warning=FALSE}
ggplot(data = movies, aes(y = audience_score, x = genre)) +
  geom_boxplot()
```

## Boxplots - axis formatting

```{r fig.height=3, fig.width=8, warning=FALSE}
ggplot(data = movies, aes(y = audience_score, x = genre)) +
  geom_boxplot() +
  theme(axis.text.x=element_text(angle = 45, hjust = 1))
```

## Density plots - border color

```{r fig.height=3, fig.width=8, warning=FALSE}
ggplot(data = movies, aes(x = runtime, color = audience_rating)) +
  geom_density() 
```

## Density plots - fill color

```{r fig.height=3, fig.width=8, warning=FALSE}
ggplot(data = movies, aes(x = runtime, fill = audience_rating)) +
  geom_density() 
```

## Density plots - fill color, with alpha

```{r fig.height=3, fig.width=8, warning=FALSE}
ggplot(data = movies, aes(x = runtime, fill = audience_rating)) +
  geom_density(alpha = 0.5) 
```

## Scatter plots

```{r fig.height=3, fig.width=8, warning=FALSE}
ggplot(data = movies, aes(x = imdb_rating, y = audience_score)) +
  geom_point(alpha = 0.5) 
```

## Smoothing

```{r fig.height=3, fig.width=8, warning=FALSE}
ggplot(data = movies, aes(x = imdb_rating, y = audience_score)) +
  geom_point(alpha = 0.5) +
  geom_smooth()
```

## Smoothing - lm

```{r fig.height=3, fig.width=8, warning=FALSE}
ggplot(data = movies, aes(x = imdb_rating, y = audience_score)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm")
```

## Barplots

```{r fig.height=3, fig.width=8, warning=FALSE}
ggplot(data = movies, aes(x = genre)) +
  geom_bar() +
  theme(axis.text.x=element_text(angle = 45, hjust = 1))
```

## Segmented barplots

```{r fig.height=3, fig.width=9, warning=FALSE}
ggplot(data = movies, aes(x = genre, fill = audience_rating)) +
  geom_bar() +
  theme(axis.text.x=element_text(angle = 45, hjust = 1))
```

## Segmented barplots - proportions

```{r fig.height=3, fig.width=9, warning=FALSE}
ggplot(data = movies, aes(x = genre, fill = audience_rating)) +
  geom_bar(position = "fill") +
  theme(axis.text.x=element_text(angle = 45, hjust = 1))
```

## Dodged barplots

```{r fig.height=3, fig.width=9, warning=FALSE}
ggplot(data = movies, aes(x = genre, fill = audience_rating)) +
  geom_bar(position = "dodge") +
  theme(axis.text.x=element_text(angle = 45, hjust = 1))
```

## Not on team `theme_grey()`?

```{r fig.height=3, fig.width=9, warning=FALSE}
ggplot(data = movies, aes(x = genre, fill = audience_rating)) +
  geom_bar(position = "dodge") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## More `ggplot2` resources

- Visit http://docs.ggplot2.org/current/ for documentation on the current version
of the `ggplot2` package. It's full of examples!

- Refer to the `ggplot2` cheatsheet: http://www.rstudio.com/wp-content/uploads/2015/12/ggplot2-cheatsheet-2.0.pdf

- Themes vignette: http://docs.ggplot2.org/dev/vignettes/themes.html

## Exercise 

Recereate the following plot. Hint: Add a `labs()` layer.

```{r echo=FALSE, fig.height=4}
ggplot(data = movies, aes(x = imdb_num_votes, y = imdb_rating, color = audience_rating)) +
  geom_point(alpha = 0.5) +
  facet_wrap(~ mpaa_rating) +
  labs(x = "IMDB number of votes", y = "IMDB rating", 
      title = "IMDB and RT scores, by MPAA rating",
      color = "Audience rating") +
  theme_bw()
```

# Shiny

## Shiny

- A web application framework for R with which you can easily turn your analyses into interactive web applications

- No HTML, CSS, or JavaScript knowledge required

## Set up

```{r eval=FALSE}
install.packages("shiny")
library(shiny)
```

- The Shiny package has built-in examples that each demonstrate how Shiny works. Each example is a self-contained Shiny app.

## Structure of a Shiny App

Shiny apps have two components:

- a user-interface script: controls the layout and appearance of your app

- a server script: contains the instructions that your computer needs to build your app

## Give it a try

```{r eval=FALSE}
runExample("01_hello")
```

## Various control widgets

<center>
http://shiny.rstudio.com/tutorial/lesson3/
</center>

## Live demo

```{r echo = FALSE}
# Define UI for application that draws a histogram
ui <- shinyUI(fluidPage(
  
  # Application title
  titlePanel("Exploring Movies"),
  
  # Sidebar with a slider input for number of bins 
  sidebarLayout(
    sidebarPanel(
      # Select variable for y-axis
      selectInput("y",
                  "Y-axis:",
                  choices = c("imdb_rating", "imdb_num_votes",
                              "critics_score", "audience_score",
                              "runtime"),
                  selected = "audience_score"
      ),
      
      # Select variable for x-axis
      selectInput("x",
                  "X-axis:",
                  choices = c("imdb_rating", "imdb_num_votes",
                              "critics_score", "audience_score",
                              "runtime"),
                  selected = "critics_score"
      ),
      
      # Select variable for color
      selectInput("z",
                  "Color by:",
                  choices = c("title_type", "genre", "mpaa_rating",
                              "critics_rating", "audience_rating")
      ),
      
      # Alpha
      sliderInput("alpha", 
                  "Alpha:", 
                  min = 0, max = 1, 
                  value = 0.5),
      
      # Horizontal line separating population and sample input
      hr(),
      
      # Sample size
      numericInput("n",
                   "Sample size:",
                   min = 1,
                   max = nrow(movies),
                   value = nrow(movies))
      
    ),
    
    # Show a plot of the generated distribution
    mainPanel(
      plotOutput("exp_plot")
    )
  )
))

# Define server logic required to draw a histogram
server <- shinyServer(function(input, output) {
  
  # create data frame for sample means to be stored
  movies_sample <- reactive({
    movies %>%
      sample_n(input$n)
  })
  
  output$exp_plot <- renderPlot({
    ggplot(data = movies_sample(), aes_string(x = input$x, y = input$y, 
                                              color = input$z)) +
      geom_point(alpha = input$alpha)
  })
})

# Run the application 
shinyApp(ui = ui, server = server)
```

# Acknowledgments

## Acknowledgments

Above materials are derived in part from the following sources:

* [RStudio Data Visualization Cheat Sheet](http://www.rstudio.com/wp-content/uploads/2015/12/ggplot2-cheatsheet-2.0.pdf)

* [RStudio Shiny Cheat Sheet](http://www.rstudio.com/wp-content/uploads/2016/01/shiny-cheatsheet.pdf)

* Package Documentation - [ggplot2](http://docs.ggplot2.org/current/), [shiny](http://shiny.rstudio.com/)

